---
title: "Linking genetic markers and crop model parameters using neural networks to enhance Genomic Prediction of integrative traits"
output: html_notebook
---


# Install and load packages, declare functions

Three libraries are needed for the purpose of this analysis. The 'here' package manages relative paths. The 'caret' package is used to create folds for the k-fold crossvalidation. As for the 'glmnet' package, it is used to perform genomic prediction using the LASSO method.
```{r}
required_packages <- c("here", "caret", "glmnet")
missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if(length(missing_packages)) install.packages(missing_packages)
invisible(lapply(required_packages, library, character.only = T))

normalizeData <- function(x, Df) {
  return((x - min(Df, na.rm=T))/(max(Df, na.rm=T) - min(Df, na.rm=T)))
}

findFold <- function(row, folds) {
  for(i in seq_along(folds)) {
    if(row %in% folds[[i]]) {
      return(i)
    }
  }
  return(NA)
}
```


# Import data

The data can be found in the 'data' sub-folder of this repository. 
The 'Markers.csv' file contains the marker data (rows, encoded as {-1, 0, 1}, see article for more information) for each genotype (columns) as well as the BLUP value of the complex integrative trait (first row of the file after the header).
The 'Parameters.csv' file contains the "observed" parameter values for each genotype (rows), for the eight considered genotypic parameters (columns). 
```{r}
marker_data <- read.table(file.path(here("data"), "Markers.csv"), sep=",", header=T)

# extract trait values from marker data
trait_data <- data.frame(Trait = as.double(marker_data[1, 2:ncol(marker_data)]))
# normalize trait data [0-1]
trait_data$Trait <- sapply(trait_data, normalizeData, Df = trait_data$Trait)
trait_data <- data.frame(Genotypes = colnames(marker_data)[2:ncol(marker_data)],
                         Trait = trait_data$Trait)

# remove row names and trait values from Markers.csv
marker_data <- marker_data[2:nrow(marker_data), 2:ncol(marker_data)] 
marker_data <- t(marker_data) # transpose dataframe so genotypes are rows and markers are cols

param_data <- read.table(file.path(here("data"), "Parameters.csv"), sep=",", header=T)
```


# Data analysis

Firstly, create the folds for the k-fold cross-validation. The folds used in the article are reported in the 'folds.csv' file of the data sub-folder in this repository.
```{r}
set.seed(1337)
k_fold <- 5
folds <- createFolds(1:nrow(trait_data), k=k_fold)

# assign fold number to each observation
trait_data$fold <- sapply(seq_len(nrow(trait_data)), findFold, folds=folds)
```

Secondly, perform k-fold cross-validation of the LASSO model and extract best lambda value
```{r}
lasso_res <- cv.glmnet(marker_data, trait_data$Trait, foldid=trait_data$fold, 
                       alpha=1, nlambda=5000, maxit=5000)
lamba_val <- lasso_res$lambda.min
```

Finally, run predictions for each fold and store results
```{r}
trait_results <- data.frame()
for(i in 1:k_fold) {
  y_test <- trait_data[folds[[i]],]
  y_train <- trait_data[-folds[[i]],]
  x_test <- marker_data[folds[[i]],]
  x_train <- marker_data[-folds[[i]],]
  
  lasso_model <- glmnet(x_train, y_train$Trait, alpha=1, lambda=lamba_val)
  lasso_predict <- predict(lasso_model, newx = x_test)
  trait_results <- rbind(trait_results, data.frame(
    Genotypes = trait_data[folds[[i]], "Genotypes"],
    Observed = trait_data[folds[[i]], "Trait"],
    Predicted = as.double(lasso_predict)
  ))
}
```

The same last two parts can be done for the parameter values
```{r}
param_results <- list()
for(p in 2:ncol(param_data)) {
  # extract parameter data and normalize
  p_data <- sapply(param_data[,p], normalizeData, Df = param_data[,p])
  
  # k-fold cross validation
  lasso_res <- cv.glmnet(marker_data, p_data, foldid=trait_data$fold, 
                       alpha=1, nlambda=5000, maxit=5000)
  
  lamba_val <- lasso_res$lambda.min
  
  # run predictions
  results <- data.frame()
  for(i in 1:k_fold) {
    y_test <- p_data[folds[[i]]]
    y_train <- p_data[-folds[[i]]]
    x_test <- marker_data[folds[[i]],]
    x_train <- marker_data[-folds[[i]],]
    
    lasso_model <- glmnet(x_train, y_train, alpha=1, lambda=lamba_val)
    lasso_predict <- predict(lasso_model, newx = x_test)
    results <- rbind(results, data.frame(
      Genotypes = param_data[folds[[i]], "Genotype"],
      Observed = p_data,
      Predicted = as.double(lasso_predict)
    ))
  }
  param_results[[colnames(param_data)[p]]] <- results
}
```